Observing and Predicting Knowledge Worker Stress, Focus and Awakeness in the Wild

### 2020-10-01 IJHCSSubmission/bp_studyMethod.tex
removing the past revision blue marks and starting with new ones for this coming revision

%!TEX root = bioPrediction_main.tex
\section{Field Study}
We conducted an eight-week field study with 14 participants using experience sampling and biometric sensors to investigate how knowledge workers experience stress \rev{and awakeness} over time and the feasibility of predicting stress, focus, and awakeness \rev{using} biometric signals.

@@

\subsection{Procedure}
We performed a study over the course of eight weeks. This study includes the collection of the participant's \rev{biometric data and self-reporting surveys during their work hours}. \rev{We also collected computer interaction data, but given privacy concerns of some participants chose not to use this data (see Section~\ref{sec:computer_interaction_data})}. We informed participants about the study purpose and procedure, handed out biometric sensors, introduced and explained the self-reporting to the participants, \rev{handed out consent forms and ensured informed consent from each participant for the study.}

After the initial setup, participants were asked to fill out three surveys each day for the following eight weeks (see Section~\ref{sec:Surveys}) \rev{and wear the biometric sensor during their work hours}. At the end of the eight weeks, we collected the biometric sensors and performed a short follow-up interview on the study and the participants' experiences.

@@

\subsubsection{Biometric Sensors}
Figure~\ref{everion} illustrates Biovotion's Everion, which we used to track the biometric signals of the study participants. The Everion is worn on the upper arm and provides continuous monitoring of certain biometric measurements.\footnote{https://biovotion.zendesk.com/hc/en-us/articles/213613165} Previous studies~\cite{zuger18,sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent} have used similar devices~\cite{Okada11,polar,fitbitCharge} to capture psycho-physiological and biometric measurements for shorter periods of time or capturing a smaller number of measurements.
\rev{A comparative study has  shown that the Everion can be used as a valid proxy for HRV metrics for knowledge workers~\cite{barrios2019a}.}

@@

Table~\ref{signals} lists the biometrics measurements that we collected using the Everion. Each measurement is collected once per second, and each recorded observation has an associated timestamp and quality rating. Data collected by the Everion are uploaded to a server, from which we downloaded the data for use in our study. \rev{We chose these biometric measurements based on previous research that has indicated their potential (see references in Table~\ref{signals}) and the availability and feasibility of collecting these measurements with low invasiveness over a long duration.}

\begin{table}[h!]
\begin{center}
\small\addtolength{\tabcolsep}{10pt}%-1pt}
\begin{tabular}{l l}
\hline
Biometric Measurement & Units of Measure \\ 
\hline
\textbf{Physical Activity} & \cite{fox1999,aldana1996}\\
\hspace{3mm}Intensity of motion & (No unit)  \\
\hspace{3mm}Energy Expenditure & Calories per second (cal/s)  \\
\hspace{3mm}Step counter & Steps  \\
\hline
\textbf{Heart} & \cite{haapalainen2010psycho,healey2005detecting,mulder1992measurement,haag2004emotion} \\
\hspace{3mm}Heart rate & Beats per Minute (bpm) \\
\hspace{3mm}Blood pulse wave & (No Unit)  \\
\hspace{3mm}Heart rate variability: RMSSD  & Milliseconds (ms) \\
%\multicolumn{2}{l}{\hspace{4mm}\textit{(RMSSD: root mean square of successive heartbeat interval differences)}} \\
\hspace{3mm}Blood oxygenation  & Percent (\%)  \\
\hspace{3mm}Blood perfusion & (No unit)  \\
\hline
\textbf{Skin} & \cite{healey2005detecting,haag2004emotion} \\
\hspace{3mm}Galvanic skin response &  kOhm  \\
\hspace{3mm}Skin temperature &  Degrees Celsius ($^{\circ}$C)  \\
\hline
\textbf{Respiration} & \cite{mulder1992measurement,healey2005detecting,haag2004emotion,masaoka1997}\\
\hspace{3mm}Respiratory rate & Breaths per Minute (bpm)  \\
\hline
%\textbf{Environmental}\\
%\hspace{3mm}Barometric pressure & Milibar (mbar)  \\
%\hline
\end{tabular}
\caption{Biometric measurements captured by the Everion, organized by category and with references to previous works using similar data. \rev{\textit{(RMSSD denotes the root mean square of successive heartbeat interval differences)}}}
\label{signals}
\end{center}
%\vspace*{-4mm}
\end{table}

\subsubsection{Computer Interaction Data}\label{sec:computer_interaction_data}
To gain a better understanding of our participants day-to-day work activities, we \rev{asked participants to install} an open source computer interaction monitor~\cite{Meyer17}. The monitor ran in the background on participants' computers and tracked the active windows, as well as the keyboard and mouse activity. Four of our participants opted out of this part of the study for privacy reasons (S6, S10, S12, and S13).
% Gail: I commented out Thomas' comment as I don't think we need to address it
%\tfC{Probably needs to state something about how it also captured non-work related browsing (see section 3.4)}
%Therefore, we only included the computer interaction data in our analysis for our more coarse-grained explanatory model but not in our analysis on the predictive model for stress in the moment.

@@

\rev{We did not ask our participants about their focus levels in the end of the day survey as focus
is more of an in-the-moment aspect than stress and awakeness.}

>>>>

%!TEX root = bioPrediction_main.tex
\section{Field Study}
We conducted an eight-week field study with 14 participants using experience sampling and biometric sensors to investigate how knowledge workers experience stress and awakeness over time and the feasibility of predicting stress, focus, and awakeness using biometric signals.

@@

\subsection{Procedure}
We performed a study over the course of eight weeks. This study includes the collection of the participant's biometric data and self-reporting surveys during their work hours. We also collected computer interaction data, but given privacy concerns of some participants chose not to use this data (see Section~\ref{sec:computer_interaction_data}). We informed participants about the study purpose and procedure, handed out biometric sensors, introduced and explained the self-reporting to the participants, handed out consent forms and ensured informed consent from each participant for the study.

After the initial setup, participants were asked to fill out three surveys each day for the following eight weeks (see Section~\ref{sec:Surveys}) and wear the biometric sensor during their work hours. At the end of the eight weeks, we collected the biometric sensors and performed a short follow-up interview on the study and the participants' experiences.

@@

\subsubsection{Biometric Sensors}
Figure~\ref{everion} illustrates Biovotion's Everion, which we used to track the biometric signals of the study participants. The Everion is worn on the upper arm and provides continuous monitoring of certain biometric measurements.\footnote{https://biovotion.zendesk.com/hc/en-us/articles/213613165} Previous studies~\cite{zuger18,sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent} have used similar devices~\cite{Okada11,polar,fitbitCharge} to capture psycho-physiological and biometric measurements for shorter periods of time or capturing a smaller number of measurements.
A comparative study has  shown that the Everion can be used as a valid proxy for HRV metrics for knowledge workers~\cite{barrios2019a}.

@@

Table~\ref{signals} lists the biometrics measurements that we collected using the Everion. Each measurement is collected once per second, and each recorded observation has an associated timestamp and quality rating. Data collected by the Everion are uploaded to a server, from which we downloaded the data for use in our study. We chose these biometric measurements based on previous research that has indicated their potential (see references in Table~\ref{signals}) and the availability and feasibility of collecting these measurements with low invasiveness over a long duration.

\begin{table}[h!]
\begin{center}
\small\addtolength{\tabcolsep}{10pt}%-1pt}
\begin{tabular}{l l}
\hline
Biometric Measurement & Units of Measure \\ 
\hline
\textbf{Physical Activity} & \cite{fox1999,aldana1996}\\
\hspace{3mm}Intensity of motion & (No unit)  \\
\hspace{3mm}Energy Expenditure & Calories per second (cal/s)  \\
\hspace{3mm}Step counter & Steps  \\
\hline
\textbf{Heart} & \cite{haapalainen2010psycho,healey2005detecting,mulder1992measurement,haag2004emotion} \\
\hspace{3mm}Heart rate & Beats per Minute (bpm) \\
\hspace{3mm}Blood pulse wave & (No Unit)  \\
\hspace{3mm}Heart rate variability: RMSSD  & Milliseconds (ms) \\
%\multicolumn{2}{l}{\hspace{4mm}\textit{(RMSSD: root mean square of successive heartbeat interval differences)}} \\
\hspace{3mm}Blood oxygenation  & Percent (\%)  \\
\hspace{3mm}Blood perfusion & (No unit)  \\
\hline
\textbf{Skin} & \cite{healey2005detecting,haag2004emotion} \\
\hspace{3mm}Galvanic skin response &  kOhm  \\
\hspace{3mm}Skin temperature &  Degrees Celsius ($^{\circ}$C)  \\
\hline
\textbf{Respiration} & \cite{mulder1992measurement,healey2005detecting,haag2004emotion,masaoka1997}\\
\hspace{3mm}Respiratory rate & Breaths per Minute (bpm)  \\
\hline
%\textbf{Environmental}\\
%\hspace{3mm}Barometric pressure & Milibar (mbar)  \\
%\hline
\end{tabular}
\caption{Biometric measurements captured by the Everion, organized by category and with references to previous works using similar data. \textit{(RMSSD denotes the root mean square of successive heartbeat interval differences)}}
\label{signals}
\end{center}
%\vspace*{-4mm}
\end{table}

\subsubsection{Computer Interaction Data}\label{sec:computer_interaction_data}
To gain a better understanding of our participants day-to-day work activities, we asked participants to install an open source computer interaction monitor~\cite{Meyer17}. The monitor ran in the background on participants' computers and tracked the active windows, as well as the keyboard and mouse activity. Four of our participants opted out of this part of the study for privacy reasons (S6, S10, S12, and S13).
% Gail: I commented out Thomas' comment as I don't think we need to address it
%\tfC{Probably needs to state something about how it also captured non-work related browsing (see section 3.4)}
%Therefore, we only included the computer interaction data in our analysis for our more coarse-grained explanatory model but not in our analysis on the predictive model for stress in the moment.

@@

We did not ask our participants about their focus levels in the end of the day survey as focus
is more of an in-the-moment aspect than stress and awakeness.

### 2020-10-01 IJHCSSubmission/bp_threats.tex
removing the past revision blue marks and starting with new ones for this coming revision

\subsection{Internal Validity}
This study tries to find correlations between
biometric features and \rev{the human aspects of} stress, focus, and awakeness.
Nonetheless, biometric signals are influenced by far more
variables than the ones this study comprehends.
Therefore, trying to draw a strong causality between the biometric
features and the aspects would be inaccurate.
To mitigate this risk,  we collected the data
in a rote environment and in a regular manner 
to minimize the number of 
external causes that may affect each participant's
biometric signals.

@@

\rev{Due to the small amount of data available to us for the purposes of this study, there is a risk
that our models may be overfitting to some degree. Such overfitting indicates that the results presented in our work may be less than optimal, however any overfitting applies strictly to the training data and given that there is no overlap between the data used for training and testing, we do not believe this invalidates our results.
We leave further optimization of our approach to future researchers.}

\subsection{Construct Validity}
A threat to the study is that
there are other factors that might either influence the
human aspects of interest or that were considered but
are unrelated biometric signals.
To mitigate this risk, we used a state-of-the-art
device that captures a large number of highly accurate biometric
signals. We collected the most commonly analyzed
biometrics that historically have shown correlation with 
\rev{the studied human aspects of stress, focus and awakeness}.
In an effort to maintain this research applicable to real-world environments, we picked \rev{the already existing Everion device}, even when, as a trade-off, we could not capture more descriptive and more intrusive signals such as SDNN, SCL, SCR, eye tracking, or brain activity.

\rev{A future more thorough statistical analysis of the relationships between
the aspects of interest, such as stress, and the physiological
data may further provide deeper insights into the data and
how it might be used in prediction.}

>>>>

\subsection{Internal Validity}
This study tries to find correlations between
biometric features and the human aspects of stress, focus, and awakeness.
Nonetheless, biometric signals are influenced by far more
variables than the ones this study comprehends.
Therefore, trying to draw a strong causality between the biometric
features and the aspects would be inaccurate.
To mitigate this risk,  we collected the data
in a rote environment and in a regular manner 
to minimize the number of 
external causes that may affect each participant's
biometric signals.

@@

Due to the small amount of data available to us for the purposes of this study, there is a risk
that our models may be overfitting to some degree. Such overfitting indicates that the results presented in our work may be less than optimal, however any overfitting applies strictly to the training data and given that there is no overlap between the data used for training and testing, we do not believe this invalidates our results.
We leave further optimization of our approach to future researchers.

\subsection{Construct Validity}
A threat to the study is that
there are other factors that might either influence the
human aspects of interest or that were considered but
are unrelated biometric signals.
To mitigate this risk, we used a state-of-the-art
device that captures a large number of highly accurate biometric
signals. We collected the most commonly analyzed
biometrics that historically have shown correlation with 
the studied human aspects of stress, focus and awakeness.
In an effort to maintain this research applicable to real-world environments, we picked the already existing Everion device, even when, as a trade-off, we could not capture more descriptive and more intrusive signals such as SDNN, SCL, SCR, eye tracking, or brain activity.

A future more thorough statistical analysis of the relationships between
the aspects of interest, such as stress, and the physiological
data may further provide deeper insights into the data and
how it might be used in prediction.

### 2020-10-01 IJHCSSubmission/elsarticle-template.tex
removing the past revision blue marks and starting with new ones for this coming revision

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{\rev{Observing and Predicting Knowledge Worker Stress, Focus and Awakeness in the Wild}}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Authors will be visible after double blind review}
%\vspace{110cm}
%\vspace{18ex}
%}

\begin{comment}

\author{Mauricio Soto}
\affiliation{%
\institution{ABB Future Labs}
\city{Raleigh}
\state{North Carolina}
}
\email{mauricio.soto@us.abb.com}

\author{Chris Satterfield}
\affiliation{%
\institution{The University of British Columbia}
\city{Vancouver}
\state{British Columbia}
}
\email{cds00@cs.ubc.ca}

\author{Thomas Fritz}
\affiliation{%
\institution{University of Zurich}
\city{Zurich}
\state{Switzerland}
}
\email{fritz@ifi.uzh.ch}

\author{Gail C. Murphy}
\affiliation{%
\institution{The University of British Columbia}
\city{Vancouver}
\state{British Columbia}
}
\email{murphy@cs.ubc.ca}

\author{David Shepherd}
\affiliation{%
\institution{Virginia Commonwealth University}
\city{Richmond}
\state{Virginia}
}
\email{shepherdd@vcu.edu}

\author{Nicholas Kraft}
\affiliation{%
\institution{UserVoice}
\city{Raleigh}
\state{North Carolina}
}
\email{kraft.nicholas.a@gmail.com} \\}

\end{comment}

\begin{abstract}
Knowledge workers face many challenges in the workplace: work is 
fragmented, disruptions are constant, tasks are complex, and work hours can be 
long. These challenges can impact knowledge workers' stress, focus and awakeness, and in turn their interaction with the digital environment, the quality of work 
performed and their productivity in general. We report on a field study with 14
knowledge workers over an eight-week period in which we investigated, using
experience sampling, how the workers experience stress, \rev{focus and awakeness} over time. During this field study, we also collected biometric data \rev{including heart- and skin-related measures}, which we then used to investigate if it is possible to predict stress, focus and awakeness, in the moment.
We observed and report on various trends in knowledge worker stress levels
over several weeks and show that machine learning models can be built
\rev{from the data of a single minimally invasive sensor}
to
predict stress, focus and awakeness \rev{accurately, with the abstract concept of focus
the hardest to predict.}
\end{abstract}

\begin{keyword}
Biometrics, \tf{Psycho-Physiological Data}, Stress, Awakeness, Focus, Computer Interaction, Ubiquitous Computing, Empirical Study, User Centered Design
\end{keyword}

>>>>

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Observing and Predicting Knowledge Worker Stress, Focus and Awakeness in the Wild}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
%\author{Authors will be visible after double blind review}
%\vspace{110cm}
%\vspace{18ex}
%}

\author{Mauricio Soto\corref{ma}}
\ead{mauricio.soto-gonzalez1@hitachi-powergrids.com}
\address{Hitachi ABB Power Grids, Raleigh, North Carolina}

\author{Chris Satterfield\corref{ma}}
\ead{cds00@cs.ubc.ca}
\address{The University of British Columbia, Vancouver, British Columbia}

\author{Thomas Fritz}
\ead{fritz@ifi.uzh.ch}
\address{University of Zurich, Zurich, Switzerland}

\author{Gail C. Murphy}
\ead{murphy@cs.ubc.ca}
\address{The University of British Columbia, Vancouver, British Columbia }

\author{David Shepherd}
\ead{shepherdd@vcu.edu}
\address{Virginia Commonwealth University, Richmond, Virginia}

\author{Nicholas Kraft}
\ead{kraft.nicholas.a@gmail.com}
\address{UserVoice, Raleigh, North Carolina}

\cortext[ma]{Main Author}

\begin{abstract}
Knowledge workers face many challenges in the workplace: work is 
fragmented, disruptions are constant, tasks are complex, and work hours can be 
long. These challenges can impact knowledge workers' stress, focus and awakeness, and in turn their interaction with the digital environment, the quality of work 
performed and their productivity in general. We report on a field study with 14
knowledge workers over an eight-week period in which we investigated, using
experience sampling, how the workers experience stress \rev{and} awakeness over time. During this field study, we also collected biometric data including heart- and skin-related measures, which we then used to investigate if it is possible to predict stress, focus and awakeness, in the moment.
We observed and report on various trends in knowledge worker stress and awakeness levels
over several weeks, finding that people tend to have certain ``baseline'' levels for these aspects and stressful days or days with low awakeness tend to cluster. We further show that machine learning models can be built
from the data of a single minimally invasive sensor
to
predict stress, focus and awakeness. Overall, we found that our models were capable of large improvements in precision and recall in comparison to a random classifier for stress (25.9\% increase over random for precision, 4.2\% for recall) and awakeness (52.4\% increase in precision, 40.8\% in recall). The abstract concept of focus proved to be
the hardest to predict (26.0\% increase in precision, 27.8\% decrease in recall).
\end{abstract}

\begin{keyword}
Biometrics, Psycho-Physiological Data, Stress, Awakeness, Focus, Computer Interaction, Ubiquitous Computing, Empirical Study, User Centered Design
\end{keyword}
